Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-08-01 surfaces-write-up/surface-write-up.tex
edits to 4.4- don't read yet

Peter: Better (on the first comment? As for the second, that might be a little tricky, because we already have $E_\mu$, so we would have to do something like $E_{i,\mu}$ (which we could of course do...)

@@

Aaron: How does $E_\mu$ depend on $i$? If you'd prefer to just say everything depends on $i$, I'm ok with it, though I still think including $i$ subscripts would be slightly
preferable. Perhaps just say near the beginning a list of all the things that will depend on $i$. I don't think you can say "everything for the remainder of the proof depends on i", because that's not true, e.g., $\ell_j$ for $j \neq i$ doesn't depend on $i$, though of course this is very confusing because you write $\ell_i$ in the definition of $\deg E$. (Won't this be more difficult than just writing a supscript or superscript $i$ though?)
\\
Peter: Where does $\ell_j$ appear?  It may be a typo... I can't remember...  I think there are only $\ell_i$'s and $\ell_{i,j}$'s.  Anyway, what I meant was every new variable defined after this point depends on $i$, not every variable used.
}
we simply write $f_i$ as a 
polynomial in $f_1, \ldots f_4$. Furthermore, these
generate all relations, since they allow us to reduce any $\prod_{j =
1}^n f_j^{c_j}$ to the canonical form where $r_i = - \lfloor d \alpha_i
\rfloor$ whenever $i > 4$.  To bound the degree of generation of these relations, we bound the degree of generation of the ideal $(z_i - \beta_i) \cap \ker(\psi)$ for each $i$.

%Note that we can view a relation from Equation
%~\ref{eqn:hirz-relations-psi} as an element of $S$ given
%by $(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)$.  
Let $\Omega\subseteq \Sigma$ be the 
set
\begin{align*}
	\Omega = & \{(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)  \in \bz^{n+1}_{\geq 0} : \\
	& c_i \ge -\alpha_i d, \text{ for all i, and } \bida_i + \sum_{j = 1}^m \bida_j c_j = \bidb_i + \sum_{j=1}^n \bidb_j c_j
= 0 \}.
\end{align*}

Let $\omega\in \Omega$.  Then we can write $\omega$ as a sum of $r_j e_j$'s for $j\in
\Te$ and $r_{j,k}e_{j,k}$ for $j\in \Tp, k\in \Tm$.  
For convenience, let $e_j$ (when it exists) be in degree $d_j$ with $j^{th}$
component $-\alpha_j d_j + \kappa_j$ for some $\kappa_j \in \bq$ and let $e_{j,k}$ (when it exists) lie in
degrees $d_{j,k}$ with $j^{th}$ component $-\alpha_j d_{j,k} + \kappa_{j,k}'$ and $k^{th}$
component $-\alpha_j d_{j,k} + \kappa_{j,k}''$ for $\kappa_{j,k}', \kappa_{j,k}'' \in \bq$.
By a careful computation, we can find $\lambda \in \Sigma$ 
and $r_{i,j} \in \bq$ (in general such a choice may be very non-unique) so that we can write
\begin{equation}\label{eqn:sigma-in-terms-of-lambda/E}
	\omega = \lambda + E
 \end{equation}

\todo{Aaron:
Does $E$ depend on $\omega$? I.e. have we fixed an $\omega$ for the rest of the proof?
If so, why can we do this?
\\
Your write, it depends on $\omega$, not just $i$.  We can do this, because we will take the maximum over all $\omega$.  What I mean by this, is for any starting relation (with initial term $\omega$ under some appropriate order) we reduce it to a different relation of the form described in the proof.  We then take the maximum of the degrees of the relation that any $\omega$ is reduced to.
}
\todo{Aaron: If $i$ has been fixed for the remainder of the proof, why 
are we using $i$ as dummy subscripts everywhere?\\
I don't follow; as in $e_i$'s?  or $e_{i,j}$'s?  $i$ is fixed in all of these statements, whereas $j$ ranges as necessary ($j$ is not fixed).}
with $E$ of the form
\begin{equation}\label{eqn:hirz-E-translation}
	E = \begin{cases}
	\frac{1}{\kappa_i} e_i  	&  \mbox{ if }i\in \Te\\
	\sum_{j\in \Tm} r_{i,j} e_{i,j}: r_{i,j}\in \bq, \sum_{j\in \Tm} r_{i,j}\kappa_{i,j}' = 1 & \mbox{ if } i\in \Tp\\
	\sum_{j\in \Tp} r_{j,i} e_{j,i}: r_{j,i} \in \bq, \sum_{j\in \Tp} r_{j,i} \kappa_{i,j}'' = 1 	& \mbox{ if } i\in \Tm\\
	\end{cases}.
\end{equation}
Specifically, if we write out $\omega = \sum_{j\in \Te} s_j e_j + \sum_{j\in \Tp, k \in \Tp} s_{j,k} e_{j,k}$, then any choice of $r_{j,k}$'s not greater than the respective $s_{j,k}$'s such that $r_{i,j} \kappa_{i,j}' =1$ (when $i\in \Tp$ but a similar statement holds for $i\in \Tm$) will suffice.  Such a choice of $r_{i,j}$'s uniquely determines the necessary $\lambda$.

From this, we can further deduce that for such an $E$,
\[
	\deg(E) \le \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in \Te \\
	\max_{j \in \Tm} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in \Tp \\
	\max_{j \in \Tp} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in \Tm \end{cases}.
\]
Let $M$ be the set of monomials terms of $\beta_i$.  Then for $\mu = \prod_{j=1}^n
{f_i}^{c_i}\in M$, we
\todo{Aaron: What is $M$? \\
Peter: I don't quite follow this question... Is your objection that ``the set of monomial terms of $\beta_i$'' is too imprecise?

Aaron: Sorry, I was trying to say that
I don't understand what ``the set of monomials in terms of beta i'' means.
Do you just mean that $M = \{\psi(\beta_1), \ldots, \psi(\beta_n)\}$?

However, I'm really baffled as to why you want to apply $\psi$ to the
$\beta_i$, since two lines later, you then apply the reverse correspondence
and want to work back in the semigroupring. Why don't you just stay in the
semigroup ring the whole time?
I think I'm misunderstanding this, but is $\omega_\mu$ just equal to
$\beta_i$ for some $i$? Also, why is $\omega_\mu$ missing a $d$ index
as its first index?
\\
Peter: $\beta_i$'s are polynomials in $f_1, \ldots f_4$.  By the monomial terms in $\beta_i$ I mean look at each monomial appearing in this polynomial and take the set of those.
}
can further compute that there exist $q_i, q_{i,j} \in \bq_{\geq 0}$ 
and $\lambda \in
\Sigma$ so that the lattice point, $\omega_\mu = (c_1 ,\ldots c_n)$,
corresponding to $\mu
$ is of the form
\[
	\omega_\mu = \lambda + E_\mu
\]
where $\lambda$ is the same as in Equation \ref{eqn:sigma-in-terms-of-lambda/E} and $E_\mu$ is defined by

\begin{equation}\label{eqn:hirz-E_mu-translation}
E_\mu = \left(\sum_{\substack{i \in \{1, 2, 3, 4\} \cap \Te}}q_i^\mu e_i\right) +
\left(\sum_{\substack{i \in \{1, 2, 3, 4\} \cap \Tp \\ j\in \Tm}} q_{i,j}^\mu
e_{i,j}\right) + \left(\sum_{\substack{i \in \{1, 2, 3, 4\} \cap \Tm \\ j\in
\Tp}} q_{j,i}^\mu e_{i,j}\right).
\end{equation}

Note that $E_\mu$ lies in the same degree as $E$ for all $\mu$.

Let $F\in R_D$ correspond to the lattice point $E$ and let $F_\mu$ correspond to the lattice point $E_\mu$.  
Then $F - \sum_{\mu \in M} F_\mu = 0$ induces a
relation in degree at most
\[
	\rho + \deg(F)
\]

\noindent
which divides $(z_i - \beta_i) \prod_{i = 1}^n y_i^{c_i}$ (recall that $\rho$, defined in Equation \ref{eqn:def-sigma}, is an upper bound for generation degree).
We can bound $\deg(F)$ (which equals $\deg(F_\mu)$ for all $\mu$) by
\[
	\deg(F) = \deg(E) \le \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in \Te \\
	\max_{j \in \Tm} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in \Tp \\
	\max_{j \in \Tp} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in \Tm \end{cases}.
\]

>>>>

Peter: Better (on the first comment)? As for the second, that might be a little tricky, because we already have $E_\mu$, so we would have to do something like $E_{i,\mu}$ (which we could of course do...)

@@

Aaron: How does $E_\mu$ depend on $i$? If you'd prefer to just say everything depends on $i$, I'm ok with it, though I still think including $i$ subscripts would be slightly
preferable. Perhaps just say near the beginning a list of all the things that will depend on $i$. I don't think you can say "everything for the remainder of the proof depends on i", because that's not true, e.g., $\ell_j$ for $j \neq i$ doesn't depend on $i$, though of course this is very confusing because you write $\ell_i$ in the definition of $\deg E$. (Won't this be more difficult than just writing a supscript or superscript $i$ though?)
\\
Peter: Where does $\ell_j$ appear?  It may be a typo... I can't remember...  I think there are only $\ell_i$'s and $\ell_{i,j}$'s.  Anyway, what I meant was every new variable defined after this point depends on $i$, not every variable used.
}
we simply write $f_i$ as a 
polynomial in $f_1, \ldots f_4$. Furthermore, these
generate all relations, since they allow us to reduce any monomial $\prod_{j =
1}^n f_j^{c_j}$ to the canonical form where $r_i = - \lfloor d \alpha_i
\rfloor$ whenever $i > 4$.  To bound the degree of generation of these relations, we bound the degree of generation of the ideal $(z_i - \beta_i) \cap \ker(\psi)$ for each $i$.

We now wish to bound the degree of generation the ideal of relations generated by these.  To do this, we first fix $i$ and a relation of the form $(f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}=0$ (note that we ignore the case $i\in \{1, 2, 3, 4\}$ as in those cases all relations listed above are just the trivial relation because $\beta_i = f_i$).  Our goal is to find a relation that the divides this, but whose degree we can bound.  We do so by considering the lattice points corresponding to the monomials appearing in the relation $(f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}$, and finding some $\lambda_\omega\in \Sigma$ that we can factor out of each monomial, to get a relation below a bounded degree.  The following argument makes this precise:

%Let $\Omega\subseteq \Sigma$ be the 
%set
%\begin{align*}
%	\Omega = & \{(d, c_1, \ldots, c_{i-1}, c_i + 1, c_{i+1}, \ldots , c_n)  \in \bz^{n+1}_{\geq 0} : \\
%	& c_i \ge -\alpha_i d, \text{ for all i, and } \bida_i + \sum_{j = 1}^m \bida_j c_j = \bidb_i + \sum_{j=1}^n \bidb_j c_j
%= 0 \}.
%\end{align*}

Consider a relation $(f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}=0$ with corresponding lattice point $\sigma: = (d, c_1, \ldots, c_{i-1}, c_{i}+1, c_{i+1}, \ldots, c_n) \in \Sigma$.
Then we can write $\sigma$ as a sum of $s_j e_j$'s for $j\in
\Te$ and $s_{j,k}e_{j,k}$ for $j\in \Tp, k\in \Tm$.  
For convenience, let $e_j$ (when it exists) be in degree $d_j$ with $j^{th}$
component $-\alpha_j d_j + \kappa_j$ for some $\kappa_j \in \bq$ and let $e_{j,k}$ (when it exists) lie in
degrees $d_{j,k}$ with $j^{th}$ component $-\alpha_j d_{j,k} + \kappa_{j,k}'$ and $k^{th}$
component $-\alpha_k d_{j,k} + \kappa_{j,k}''$ for $\kappa_{j,k}', \kappa_{j,k}'' \in \bq$.

We note that for $(f_i - \beta_i)\prod_{j=1}^n f_j^{c_j}=0$ to be a relation, every monomial must actually lie in $R_D$; this implies that $s_i\kappa_i\ge 1$, $\sum_{j\in \Tm} s_{i,j}\kappa_{i,j}' \ge 1$, and $\sum_{j\in \Tp} s_{i,j}\kappa_{j,i}'' \ge 1$ if $i\in \Te$, $i\in \Tp$, and $i\in \Tm$ respectively.

If $i\in \Tp$ choose $r_{i,j}\in \bq_{\ge 0}$ for all $j\in \Tm$ such that $\sum_{j\in \Tm} r_{i,j}\kappa_{i,j}' = 1$; similarly, if $i\in \Tm$ choose $r_{j,i}\in \bq_{\ge 0}$ for all $j\in \Tp$ such that $\sum_{j\in \Tm} r_{j,i}\kappa_{j,i}'' = 1$.
Define $E$ by
\begin{equation}\label{eqn:hirz-E-translation}
	E := \begin{cases}
	\frac{1}{\kappa_i} e_i  &  \mbox{ if }i\in \Te\\
	\sum_{j\in \Tm} r_{i,j} e_{i,j} & \mbox{ if } i\in \Tp\\
	\sum_{j\in \Tp} r_{j,i} e_{j,i} 	& \mbox{ if } i\in \Tm.
	\end{cases}
\end{equation}
Define
\begin{equation}\label{eqn:defn-lambda}
	\lambda = \sigma - E\in \Sigma.
 \end{equation}
Note that
\[
	\deg(E) \le \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in \Te \\
	\max_{j \in \Tm} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in \Tp \\
	\max_{j \in \Tp} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in \Tm. \end{cases}
\]

Let $M_i$ be the set of monomials terms of $\beta_i$ (recall $\beta_i$ is a polynomial in $f_1, \ldots f_4$).  Let $\mu = \prod_{j=1}^4
{f_j}^{h_j}\in M_i$ and consider the lattice point (in the same degree $d$ as $\sigma$)
\[
	\sigma_\mu = (d, c_1 + h_1, \ldots, c_4+h_4, c_5, \ldots, c_n).
\]
Define 
\[
	E_\mu = \sigma_\mu - \lambda.
\]
A quick computation deduces that $E_\mu$ lies in $\Sigma$ (using the condition on the $r_{i,j}$'s and $r_{j,i}'s$) and has the same degree as $E$.

Let $F\in R_D$ be the monomial given by the lattice point $E$ and $F_\mu \in R_D$ be the monomial given by the lattice point $E_\mu$ for each $\mu \in M_i$.  
Then $F - \sum_{\mu \in M} F_\mu = 0$ is a relation %in degree
%\[
%	\deg(F) = \deg(E)
%\]
which divides $(z_i - \beta_i) \prod_{i = 1}^n y_i^{c_i}$. 
We can bound $\deg(F)$ (which equals $\deg(F_\mu)$ for all $\mu$) by
\[
	\deg(F) = \deg(E) \le \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in \Te \\
	\max_{j \in \Tm} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in \Tp \\
	\max_{j \in \Tp} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in \Tm. \end{cases}
	\]
